Validating risk scores

### 2020-10-22 manuscript/competing_risks.Rmd
Work on manuscript

```{r}
hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
```

>>>>

```{r}
hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
write_csv(hypothesis_res_all, path = here("manuscript", "/competing_risks_res.csv"))
```

### 2020-10-22 manuscript/manuscript.Rmd
Work on manuscript

---
title: "Detailed disease progression of 97 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  bookdown::pdf_document2:
     toc: false
  # bookdown::word_document2: 
  #     toc: false
  # bookdown::pdf_book: 
  #    base_format: rticles::arxiv_article
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
bibliography: Covid19.json
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
---

@@

devtools::load_all()
library(tidyverse)
library(patchwork)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide
```

@@

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8] where an underlying continous process with given rates only between the neighbouring state and the transition matrix across all states is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the unobserved state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We primarily use a model that explicitly uses all of the recorded levels of breathing support and has a hidden (unobserved) element whether the patient is improving or worsening. The states and available transitions are shown in Figure&nbsp;\@ref(fig:modelstates)b.

```{r modelstates, fig.cap="States used in the competing risk model (a) and in the hidden Markov model (b). AA = Ambient air, Oxygen = Nasal oxygen, NIPPV = Non-invasive positive-pressure ventilation, ECMO = Extra-corporeal membrane oxygenation. In both models the 'Death' and 'Discharged' states are terminal. In the hidden Markov model, the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
knitr::include_graphics(here::here("manuscript","static_figs","model_states.pdf"))
```

@@

@@

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e. where we have gathered all the input features). This resulted in five usable models: [@http://zotero.org/users/5567156/items/EN4GICJ6; @http://zotero.org/users/5567156/items/INZVABQI; @http://zotero.org/users/5567156/items/3A2RN4ZS; @http://zotero.org/users/5567156/items/4DDUVEKS; @http://zotero.org/users/5567156/items/X4GWJIP5].

@@

```{r}
n_patients <- nrow(data$patient_data)
n_hospitals <- length(unique(data$patient_data$hospital_id))
```

@@

```{r}
bayesian_cols <- cols(
  hypothesis = col_character(),
  model = col_character(),
  adjusted = col_character(),
  estimand = col_character(),
  widest_CI_excl_reference = col_double(),
  point_estimate = col_double(),
  ci_low = col_double(),
  ci_high = col_double()
)
res_bayesian <- 
  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
        )

@@

>>>>

---
title: "Detailed disease progression of 191 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
     toc: false
  # bookdown::pdf_book: 
  #    base_format: rticles::arxiv_article
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
bibliography: Covid19.json
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
---

@@

devtools::load_all()
library(tidyverse)
library(patchwork)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide

n_patients <- nrow(data$patient_data)
n_hospitals <- length(unique(data$patient_data$hospital_id))
```

@@

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighbouring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c). The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction.

```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c). AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation. In all models the 'Death' and 'Discharged' states are terminal. In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
```

@@

Finally we used a set of Bayesian regression models implemented with the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. Those included overall survival, state at day 7 or 28 as either binary or ordinal outcome, Bayesian version of the Cox proportional-hazards model and time-to-event models.

@@

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e. where we have gathered all the input features). This resulted in five usable models: [@http://zotero.org/users/5567156/items/EN4GICJ6; @http://zotero.org/users/5567156/items/INZVABQI; @http://zotero.org/users/5567156/items/3A2RN4ZS; @http://zotero.org/users/5567156/items/4DDUVEKS; @http://zotero.org/users/5567156/items/X4GWJIP5]. We used

@@

@@

```{r}
bayesian_cols <- cols(
  hypothesis = col_character(),
  model = col_character(),
  adjusted = col_character(),
  estimand = col_character(),
  widest_CI_excl_reference = col_double(),
  point_estimate = col_double(),
  ci_low = col_double(),
  ci_high = col_double()
)
res_bayesian <- 
  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
        read_csv(here::here("manuscript", "analysis_brms.csv"), col_types = bayesian_cols),
        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
        )

@@

```{r}
plot_multiverse(all_hypo %>% filter(group == "favipiravir"))
```

### 2020-10-22 manuscript/risk_scores.Rmd
Work on manuscript

bello_chavolla_data <- at_admission %>%
  inner_join(atb_first_day, by = "patient_id") %>%
  mutate(bc_base =  3 *  (age >= 65) +
           diabetes +
           5 * (diabetes & age < 40) +
           (-6) * (age < 40) +
           is_obese +
           3 * renal_disease +
           COPD,
         bc0 = bc_base + 7 * atb0,
         bc1 = bc_base + 7 * atb1
  )

@@

```{r, fig.cap = "The AUC and it's 95% credible intervals for the Bello-Chavolla et al. score. The thick blue line marks AUC of 0.5, i.e. the point where reverting the score would yield better predictions. The thin gray dashed line marks the AUC from the original study for easier comparison. We use 3 variants of the score - `base` means not inferring pneumonia at all, `day0` means that pneumonia is inferred from antibiotics at hospitalization, `day1` - inferring pneumonia from antibiotics within 1 day of hospitalization "}
bc_auc <- rbind(
  my_auc_summary(alive_30 ~ bc_base, bello_chavolla_data, note = "base"),
  my_auc_summary(alive_12 ~ bc_base, bello_chavolla_data, note = "base"),
  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data, note = "day0"),
  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data, note = "day0"),
  my_auc_summary(alive_30 ~ bc1, bello_chavolla_data, note = "day1"),
  my_auc_summary(alive_12 ~ bc1, bello_chavolla_data, note = "day1")
)

>>>>

bello_chavolla_data <- at_admission %>%
  inner_join(atb_first_day, by = "patient_id") %>%
  mutate(bc_base =  3 *  (age >= 65) +
           diabetes +
           5 * (diabetes & age < 40) +
           (-6) * (age < 40) +
           is_obese +
           3 * renal_disease +
           COPD,
         bc0 = bc_base + 7 * (atb0 & admitted_for_covid),
         bc1 = bc_base + 7 * (atb1 & admitted_for_covid)
  )

@@

```{r, fig.cap = "The AUC and it's 95% credible intervals for the Bello-Chavolla et al. score. The thick blue line marks AUC of 0.5, i.e. the point where reverting the score would yield better predictions. The thin gray dashed line marks the AUC from the original study for easier comparison. We use 3 variants of the score - `base` means not inferring pneumonia at all, `day0` means that pneumonia is inferred from being admitted for covid and antibiotics at hospitalization, `day1` - inferring pneumonia from antibiotics within 1 day of hospitalization "}
bc_auc <- rbind(
  my_auc_summary(alive_30 ~ bc_base, bello_chavolla_data, note = "base"),
  my_auc_summary(alive_12 ~ bc_base, bello_chavolla_data, note = "base"),
  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data, note = "day0"),
  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data, note = "day0"),
  my_auc_summary(alive_30 ~ bc1, bello_chavolla_data, note = "day1"),
  my_auc_summary(alive_12 ~ bc1, bello_chavolla_data, note = "day1")
)
